The variability of Betelgeuse explained by surface convection\thanks{Based on observations obtained at the T\'elescope Bernard Lyot (TBL) at Observatoire du Pic du Midi, CNRS/INSU and Universit\'e de Toulouse, France.}

### 2024-06-04 main.tex
Correct IRAP affiliation
Last modif

   \institute{IRAP, Universit\'e de Toulouse, CNRS, CNES, UPS.  14, Av. E. Belin. 31400 Toulouse, France 
             }

@@

  \abstract
 % context heading (optional)
  {Betelgeuse is a red supergiant (RSG) that is known to vary semi-regularly on both short and long timescales.}
  % {} leave it empty if necessary  
   {The origin of the short period of Betelgeuse has often been associated to radial pulsations but could also be due to the convection motions present at the surface of RSGs. Here, we investigate the link between surface activity and the variability of the star.   }
  % aims heading (mandatory)
   {Linear polarization in Betelgeuse is a proxy of convection which is unrelated to pulsations. From the 10 years of spectropolarimetric data of Betelgeuse, we seek for periodicities in the least-squares deconvolution (LSD) profiles of Stokes $I$, $Q$, $U$ and the total linear polarization using the Lomb-Scargle periodogram. }
  % methods heading (mandatory)
   {We find that the 400 d period is present in the LSD profile of Stokes $I$ but is close to a peak of the window function of the TBL. Two periods of 330 d and 200 d are present in the periodogram of Stokes $Q$ and $U$, showing that we recover the variability of Betelgeuse using surface convection. We also speculate that change in variability of Betelgeuse after the Great Dimming is due to the non-equilibrium of the photosphere, destroying the large convective cells and leading to shorter timescales in the light curve of the star. }
     % results heading (mandatory)
   {}
  % conclusions heading (optional), leave it empty if necessary

>>>>

   \institute{
            Institut de Recherche en Astrophysique et Plan\'etologie,
            Universit\'e de Toulouse, CNRS, IRAP/UMR 5277,
            14 avenue Edouard Belin, F-31400, Toulouse, France  
             }

@@

  \abstract
 % context heading (optional)
  {Betelgeuse is a red supergiant (RSG) that is known to vary semi-regularly on both short and long timescales. The origin of the short period of Betelgeuse has often been associated to radial pulsations but could also be due to the convection motions present at the surface of RSGs.}
  % {} leave it empty if necessary  
   { We investigate the link between surface activity and the variability of the star, ignoring pulsations.   }
  % aims heading (mandatory)
   {Linear polarization in Betelgeuse is a proxy of convection which is unrelated to pulsations. Using  10 years of spectropolarimetric data of Betelgeuse, we seek for periodicities in the least-squares deconvolution (LSD) profiles of Stokes $I$, $Q$, $U$ and the total linear polarization using  Lomb-Scargle periodograms. }
  % methods heading (mandatory)
   {We find similar periods in linear polarization signals than in photometric variability.
   The 400 d period is too close to a peak of the window function of our data. But the two periods of 330 d and 200 d are present in the periodogram of Stokes $Q$ and $U$, showing that  the variability of Betelgeuse can be interpreted as due to  surface convection. }
     % results heading (mandatory)
   {Since linear polarization in the spectrum of Betelgeuse is not known to vary with pulsations, but is linked to surface convection, and 
   since similar periods are found in time series of photometric measurements and spectropolarimetry, we conclude that the photometric 
   variability is due to the surface convective structures, and not to any pulsation phenomenon.}
  % conclusions heading (optional), leave it empty if necessary
